Reanalysis Step 2: Multiple Imputation

### 2025-09-01 reanalysis-2-multiple_imputation.qmd
mi copy fix

@@

- Mode: `r analysis_mode`
- Imputations: `r n_imputations`
- Iterations: `r n_iterations`
- Estimated time: `r time_estimate`

```{r mice-run}
log_message("Starting MICE imputation...")
imp_start <- Sys.time()

@@

@@

@@

### Completeness Check

>>>>

# CACHING CONFIGURATION FOR MICE
# Set to TRUE to use cached MICE results (skip imputation, only run post-processing)
use_cached_mice <- TRUE
mice_cache_file <- file.path(reanalysis_data_dir, "mice_results.rds")
log_message(paste0("MICE caching: ", ifelse(use_cached_mice, "ENABLED", "DISABLED")))

@@

- Mode: `r analysis_mode`
- Imputations: `r n_imputations`
- Iterations: `r n_iterations`
- Estimated time: `r time_estimate`
- MICE caching: `r ifelse(use_cached_mice, "ENABLED", "DISABLED")`
- Cache file: `r if(use_cached_mice && file.exists(mice_cache_file)) "EXISTS (will load)" else "NOT FOUND (will run fresh)"`

```{r mice-run}
# Check for cached MICE results
if (use_cached_mice && file.exists(mice_cache_file)) {
  log_message("Loading cached MICE results...")
  log_message(paste0("Cache file: ", mice_cache_file))

  mice_results <- readRDS(mice_cache_file)

  # Verify cached results match expected parameters
  cached_m <- mice_results$m
  cached_maxit <- mice_results$iteration

  log_message(paste0("Loaded ", cached_m, " imputations from cache"))
  log_message(paste0("Original run used ", cached_maxit, " iterations"))

  if (cached_m != n_imputations) {
    warning(paste0("Cached imputations (", cached_m, ") differ from requested (", n_imputations, ")"))
  }

  imp_time <- 0  # No time spent on imputation
  log_message("Using cached MICE results - skipping imputation step")

} else {
  # Run fresh MICE imputation
  if (use_cached_mice && !file.exists(mice_cache_file)) {
    log_message("Cache file not found - running fresh imputation")
  } else {
    log_message("Cache disabled - running fresh imputation")
  }

  log_message("Starting MICE imputation...")
  imp_start <- Sys.time()

@@

  # Save to cache for future use
  saveRDS(mice_results, mice_cache_file)
  log_message(paste0("Saved MICE results to cache: ", mice_cache_file))
}

@@

  # CRITICAL: Copy person_dr_max (imputed) to person_dr for create_dr_severity_variables
  # person_dr_max is the person-level DR that was imputed
  # person_dr needs to be set for each eye from person_dr_max
  if ("person_dr_max" %in% names(imp_data)) {
    # Convert from factor if needed
    if (is.factor(imp_data$person_dr_max)) {
      # Convert factor levels (No_DR=1, NPDR=2, PDR=3) to numeric (0, 1, 2)
      imp_data$person_dr <- as.numeric(imp_data$person_dr_max) - 1
    } else {
      imp_data$person_dr <- imp_data$person_dr_max
    }
  }

@@

### Comprehensive Completeness Check

```{r comprehensive-completeness}
# Define key variables that should have been imputed
key_imputed_vars <- c(
  # Visual acuity variables
  "baseline_VA_logMAR",
  "intermediate_VA_logMAR", 
  "outcome_VA_logMAR",

  # DR severity (derived from person_dr)
  "person_dr",
  "dr_severity",

  # Comorbidity indices
  "CCI",
  "DCSI",

  # Treatment variables
  "catsurg_before_hitplus_cat",
  "glaucoma_bef_hitplus_cat",
  "otherretina_bef_hitplus_cat"
)

# Check completeness for each imputation
completeness_summary <- lapply(1:min(5, n_imputations), function(i) {
  imp_data <- imputed_datasets[[i]]

  # Check each key variable
  var_status <- lapply(key_imputed_vars, function(v) {
    if (v %in% names(imp_data)) {
      n_missing <- sum(is.na(imp_data[[v]]))
      pct_missing <- round(100 * n_missing / nrow(imp_data), 2)
      data.frame(
        Variable = v,
        Imputation = i,
        N_Missing = n_missing,
        Pct_Missing = pct_missing,
        Status = ifelse(n_missing == 0, "Complete", paste0(n_missing, " missing"))
      )
    } else {
      data.frame(
        Variable = v,
        Imputation = i,
        N_Missing = NA,
        Pct_Missing = NA,
        Status = "Variable not found"
      )
    }
  })

  do.call(rbind, var_status)
})

# Combine all results
completeness_df <- do.call(rbind, completeness_summary)

# Create summary table by variable across imputations
completeness_wide <- completeness_df |>
  select(Variable, Imputation, N_Missing) |>
  pivot_wider(names_from = Imputation, 
              values_from = N_Missing,
              names_prefix = "Imp_")

kable(completeness_wide, 
      caption = "Missing values by variable across first 5 imputations (should all be 0)",
      align = c('l', rep('r', min(5, n_imputations))))

# Flag any variables with remaining missing data
problem_vars <- completeness_df |>
  filter(N_Missing > 0) |>
  select(Variable, Imputation, N_Missing) |>
  unique()

if (nrow(problem_vars) > 0) {
  cat("\n⚠️ WARNING: The following variables still have missing data after imputation:\n")
  kable(problem_vars)
} else {
  cat("\n✓ All key variables are complete across all checked imputations\n")
}

# Compare before and after imputation
cat("\n### Before vs After Imputation Summary\n")

original_missing <- mice_results$nmis[key_imputed_vars[key_imputed_vars %in% names(mice_results$nmis)]]
post_imp_missing <- completeness_df |>
  filter(Imputation == 1) |>
  group_by(Variable) |>
  summarise(Post_Missing = sum(N_Missing, na.rm = TRUE)) |>
  filter(Variable %in% names(original_missing))

comparison_df <- data.frame(
  Variable = names(original_missing),
  Original_Missing = original_missing,
  Post_Imputation_Missing = 0  # Will be updated if any found
)

# Update with actual post-imputation values if available
for (v in comparison_df$Variable) {
  if (v %in% post_imp_missing$Variable) {
    comparison_df[comparison_df$Variable == v, "Post_Imputation_Missing"] <- 
      post_imp_missing[post_imp_missing$Variable == v, "Post_Missing"]
  }
}

comparison_df$Status <- ifelse(comparison_df$Post_Imputation_Missing == 0, 
                               "✓ Complete", 
                               "⚠️ Still missing")

kable(comparison_df,
      caption = "Imputation effectiveness: Original vs Post-imputation missing counts",
      col.names = c("Variable", "Original Missing", "After Imputation", "Status"),
      row.names = FALSE)
```
